Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-05-13 paper/paper.tex
Papper edits

\begin{document}
\title{Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales}
\author{
Victoir Veibell\footnote{vveibell@gmu.edu}
\and
R.S. Weigel\footnote{rweigel@gmu.edu}}

@@

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
This paper compares various magnetosphere conditions around the onset of geomagnetic events, defined as decreases in $D_{st}$ below a threshold value or $\rho_{eq}$ above a threshold value.

\vspace{2em}
\end{abstract}
 \end{@twocolumnfalse}
  ]

@@

The dependence of the plasma trough density on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of electron number density found a significant dependence on the plasmasphere boundary on geomagnetic activity, but there the relationship between geomagnetic activty and density in the plasma trough (\cite{Chappell1970}; \cite{Anderson1993}) was not explicitly considered.

\cite{Lotoaniu1999PlasmaMassDensity} used measurements from the Plasma Wave Experiment instrument on the CRRES satellite and compared in-situ measured plasma mass density to that inferred by ground-based ULF wave measurements from the CANOPUS array. By extending the ground footprints to the plasmatrough, they found agreement between the two methods for certain plasmapause conditions.  \vinote{More details}

@@

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) \inote{With what?}.  Although the correlation in the plasma trough was not calculated, the data presented \inote{for X events} show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

@@

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 dependences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northward and southward IMF.

@@

To determine if the magnitude of the density medians on a given day were statistically different than another day, we used a two-population t-test.  For the interval of 1989-1991, corresponding to the top panel of \ref{DailyAveragedDstEvents}, the null hypothesis that any of the medians from days -1, 0, and 1 are different from each other is rejected with a maximimum p value of \vnote{??}.

@@

Both of the results in \ref{DailyAveragedDstEvents} indicate that there is not a statictically significant change in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of \ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as \ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of \ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slighly elevated at the time of the threhshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.  At this timescale no obvious change in mass density at hourly timescale is visible. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation because there is no obvious connection on a short timescale.  One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long-term correlation of density and $D_{st}$.

Results of test on different between means of $\rho_{eq}$ shown in the bottom panel of \ref{DailyAverage} on days near or on the day of a $D_{st}$ event.

It was found that $\rho_{eq}$ enhancements do not clearly correspond to a signature in $D_{st}$, but that there is a trend for such enhancements to occur when $B_{z}$ in the IMF is positive.  This is consistent with \cite{Takahashi2010} who note that very large increases in $\rho_{eq}$ often corresponded to movement of the plasmasphere to beyond geosynchronous distances.

This method finds 669 periods of $D_{st}$ crossing beneath the selected threshold between May 1983 and August 1991 with an average duration of 9 hours and a median duration of 3 hours \vinote{Make sure I defined "duration"}. Figure \ref{HourlyAveragedDstEvents} shows the average values of all events over a window of 24 hours before onset and 48 hours after. Figure \ref{HourlyAveragedDstEvents}a shows events selected by looking for event onsets where $D_{st}$ crossed a threshold of $-40$~nT. The dashed red lines indicate plus and minus one standard deviation of values from all events that were used to compute the median. The final plot in the stack shows both $\rho_{eq}$ and a bar plot of how many valid data points went into the $\rho_{eq}$ average. Since $\rho_{eq}$ comes from a sparser dataset, it has less valid points contributing to the averages than the other parameters in the stack.

Figure \ref{HourlyAveragedDstEvents} shows a definite spike in the $Z$ component of the magnetic field, as well as the defined drop in $D_{st}$, but no obvious change in mass density at an hourly timescale. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation since there's no obvious connection on a short timescale. One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long term correlation of density and $D_{st}$.

The bottom panel of \ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.

@@

An alterntive type of event can be defined that corresponds to large increases in $\rho\ $.  In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event.  Here we see the expected result that such events tend to be associated with positive $B_z^{IMF}$.  To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours.

Figure~\label{fig:RhoBinned}(a) shows the 
The epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically signifiant difference.

\section{F10.7 dependence}
In an effort to analyze the dependence of $\rho_{eq}$ on $F_{10.7}$, a few tests were performed. \cite{Takahashi2010} mention a strong correlation between the two. The long term correlation could be a bias for $D_{st}$'s effects, so a linear model was created, recreating mass density purely from $F_{10.7}$ in the form of $\rho_{eq}(t)=a\cdot F_{10.7}(t)$. This re-created $\rho_{eq}$ shows around a 45\% correlation with the actual $\rho_{eq}$, suggesting a strong influence, while doing the same procedure with $D_{st}$ shows only a 20-25\% correlation.

Figure \ref{f107bin} takes the events and bins them by the value of $F_{10.7}$ at event onset, then looks at both $\rho_{eq}$ and $B_z$. Error bars calculated in the same manner as before are added to the highest and lowest bins. A significant difference is seen in $\rho_{eq}$ between events taking place during periods of high $F_{10.7}$ and low $F_{10.7}$, while a much less significant, but still noticeable, difference is seen in $B_z$. Also of interest is how $\rho_{eq}$ only sharply reacts to $D_{st}$ events during periods of high $F_{10.7}$. The positive relationship seen in \cite{Takahashi2010} could be due to the short time period looked at, and that it coincides with periods of higher $F_{10.7}$.

Looking at GOES 2,5,6, and 7 shows that this post-onset $D_{st}$ spike happens in almost every case for periods of high $F_{10.7}$. Creating a linear model directly relating variables (and combinations of variables) to $\rho_{eq}$ also shows that $F_{10.7}$ most often has the highest correlation with $\rho_{eq}$. A sample of these correlations is shown in Table \ref{perltable}. The correlations were made by training a one-input linear model of 5-hour post-onset medians of each variable (or variables) against the median of $\rho_{eq}$ at, and four hours after, event onset. This was done by randomly sampling half of the data, generating the model, and testing against the other half, doing this process 100 times and then taking the median correlations.

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events in the interval 1983-1991; compare to bottom panel {DailyAveragedDstEvents}. Bottom: Same as Top except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.}
\label{HourlyAveragedDstEvents}
\end{figure}

>>>>

\begin{document}
\title{Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit}
\author{
Victoir Veibell\footnote{vveibell@gmu.edu}
\and
R.S. Weigel\footnote{rweigel@gmu.edu}}

@@

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
In this work we consider two types of events, determined by decreases in $D_{st}$ below a threshold value or increases in the $\rho_{eq}$ above a threshold value on 1-hour and 1-day time scales using the \cite{Denton2007} dataset.  From the $D_{st}$ events, we find that there is no statistically significant relationship between either solar wind or geomagnetic activity under conditions of strong geomagnetic activity when hourly or daily averages of $\rho_{eq}$ are considered.  Although very large geomagnetic storms have been observed to correspond to large increases in the plasma trough mass density, in general $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or the solar wind parameters $B_{z}$ and $P_{dyn}$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.
\vspace{2em}
\end{abstract}
\hrule
 \end{@twocolumnfalse}
  ]

@@

The dependence of the plasma trough density on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of electron number density found a significant dependence on the plasmasphere boundary on geomagnetic activity (\cite{Chappell1970}; \cite{Anderson1993}), but there the relationship between geomagnetic activity and density in the plasma trough was not explicitly considered.

@@

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation in the plasma trough was not calculated, the data presented for four events show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

@@

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 dependence.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northward and southward IMF.

@@

To determine if the magnitude of the density medians on a given day were statistically different than another day, we used a two-population t-test.  For the interval of 1989-1991, corresponding to the top panel of \ref{DailyAveragedDstEvents}, the null hypothesis that any of the medians from days -1, 0, and 1 are different from each other is rejected with a maximum p value of. The results are shown in Table~\ref{Table1}.

These tests indicate that there is not a statistically significant change in medians of $\rho_{eq}$ on the day before and day after a threshold crossing in $D_{st}$.

@@

Because the dynamics of the magnetopshere take place on a time scale of less than one day, we have also considered hourly averages as shown in the top panel of Fig.~\ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as Fig.~\ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of Fig.~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small and in general not statistically significant.  At this timescale no obvious change in mass density at hourly timescale is visible. This points to an issue with only looking at long-timescale trends between density and $D_{st}$, and allows for the possibility that other factors are influencing the long term correlation because there is no obvious connection on a short timescale.  One possibility is that, as suggested in \cite{Takahashi2010}, $F_{10.7}$ plays a significant role in driving long term density values which biases the long-term correlation of density and $D_{st}$.

The bottom panel of Fig.~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there were an average of \inote{X} samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010} and the observation that solar wind parameters have very little correlation independent of $F_{10.7}$ (\cite{Denton2016}).  To determine of this relationship can be observed in the events of the top panel of Fig.~\ref{HourlyAveragedDstEvents}, we sorted the events by the average $D_{st}$ values in the time interval of 0-12 hours, split the events in two and recomputed the epoch averages of $\rho_{eq}$.  The result is that there is no statistically significant difference in the mass density epoch averages for large versus small storms.  This indicates that to confirm the $\rho_{eq}$ dependence on large geomagnetic storms would require a dataset with a much larger number of strong geomagnetic storms.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho\ $.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Here we see that such events tend to be associated with positive $B_z^{IMF}$ and low geomagnetic activity (\cite{Denton2016}).

To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours.

Figure~\label{fig:RhoBinned}(a) shows the epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically significant difference.

\subsection{Conclusions}

@@

\begin{figure}[tp!]
\centering
\includegraphics[scale=0.40]{figures/stormavs-dst.eps}
\rule[1ex]{5cm}{1pt}
\includegraphics[scale=0.40]{figures/stormavs-dd12-GOES6.eps}
\caption{$D_{st}$ events from GOES 6 using hourly medians. Top: Events in the interval 1983-1991; compare to bottom panel Fig.~\ref{DailyAveragedDstEvents}. Bottom: Same as Top except for constraint that $D_{st}$ stayed below $-50$~nT for at least 12~hours after crossing below $-50$~nT.}
\label{HourlyAveragedDstEvents}
\end{figure}
